---
title: "MT Cars Analysis"
author: "Jayesh Gokhale"
date: "5/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on Motor Trend Cars Dataset

## Executive Summary

The data set is based on 1974 Motor Trend US magazine and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles.

We have to answer two questions - 

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions.

Numeric variables - **mpg (Miles/(US) gallon)**, disp (Displacement), hp (Gross Horsepower), drat (Gear Axle Ratio), wt (Weight in 1000 lbs) and qsec (1/4 mile time) 

Categorical Variables - cyl (Number of cylinders), vs (Engine: 0 = v-shaped, 1 = straight), **am (Transmission: 0 = automatic, 1 = manual)**, gear (Number of forward gears), carb (Number of Carburetors)

```{r loadLibraries, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(ggplot2)
library(RColorBrewer)
library(knitr)
library(kableExtra)
library(reshape2)
library(ggpubr)
```

```{r mutateData, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
mt <- mtcars
mt$am <- as.factor(ifelse(mt$am==0,"automatic","manual"))
mt$vs <- as.factor(ifelse(mt$vs==0,"v-shaped","straight"))
mt$cyl <- as.factor(mt$cyl)
mt$gear <- as.factor(mt$gear)
mt$carb <- as.factor(mt$carb)
```

```{r basicboxplot, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
getBoxPlot <- function(df,xvar,yvar,fillvar,xlabel,ylabel,filllabel)
{
  mybox <- ggplot(data=df, aes(x=xvar,y=yvar,fill=fillvar)) +
  geom_boxplot(outlier.colour="black", 
               outlier.size=2,position=position_dodge(1)) + 
  labs(title = "",
       subtitle = "",
       y = ylabel, x = xlabel) +
    scale_fill_brewer(name = filllabel,palette="Dark2")  +
  theme(plot.title = element_text(hjust = 0.5)) 
  
  return(mybox)
}

plotVS <- getBoxPlot(mt,mt$vs,mt$mpg,mt$am,"Engine","MPG","Transmission")
plotCYL <- getBoxPlot(mt,mt$cyl,mt$mpg,mt$am,"No of Cylinders","MPG","Transmission")
plotGEAR <- getBoxPlot(mt,mt$gear,mt$mpg,mt$am,"No of forward gears","MPG","Transmission")
plotCARB <- getBoxPlot(mt,mt$carb,mt$mpg,mt$am,"No of carburetors","MPG","Transmission")
```

## BoxPlot Observations (Box Plot in Appendix)

1. Manual transmission does appear to have significantly higher MPG across Engines, Carburetors and No of forward gears wherever applicable
2. For 4 cylinder engines, manual transmission seems to give higher MPG
3. **In 6 & 8 cylinder engines, it seems to be too close to call.**

```{r heatmapCode, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
correlation.matrix <- cor(mt[,c("mpg","disp","hp","drat","wt","qsec")])
correlation.matrix <- round(correlation.matrix,2)
correlation.matrix[upper.tri(correlation.matrix)] <- NA
diag(correlation.matrix) <- NA
row.corr.matrix <- melt(correlation.matrix,na.rm=TRUE)
colnames(row.corr.matrix) <- c("F1","F2","CORR")
heatMap <- ggplot(data = row.corr.matrix, aes(x=F1, y=F2, fill=CORR)) + 
  geom_tile(color="white")+
  labs(title = "Correlation Coefficient Heat Map",
       subtitle = "",
       y = "Features", x = "Features") + 
  scale_fill_gradient2(low = "red", high = "darkgreen", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Correlation\nCoefficient") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()+
  geom_text(aes(x=F1, y=F2, label = CORR), color = "black", size = 4)
```

## Heatmap Observations (Heatmap in Appendix)
1. We can see that wt, disp and hp are negatively highly correlated to mpg
2. drat is positively correlated to mpg. 
3. We can also observe that wt, disp, hp and drat are correlated amongst themselves (either + or -). 

Let us now look at the relationship between am (Transmission) and mpg while adjusting for each of the four numeric variables viz. wt, disp, hp and drat. The adjustment Plot is in the appendix.

```{r adjustmentplotCode, echo=FALSE, warning=FALSE, message=FALSE}
getAdjPlot <- function(df,xvar,yvar,fillvar,xlabel,ylabel,fillabel)
{
  adjPlot <- ggplot(data=df, aes(x=xvar,y=yvar,fill=fillvar)) +
  geom_point(aes(color=fillvar)) + geom_smooth(method=lm) + 
  labs(title = "",
       subtitle = "",
       y = ylabel, x = xlabel) +
  scale_fill_brewer(name = fillabel,palette="OrRd")  +
  theme(plot.title = element_text(hjust = 0.5)) 
  
  return(adjPlot)
}

plotWeight <- getAdjPlot(mt,mt$wt,mt$mpg,mt$am,"Weight","MPG","Transmission")
plotDisp <- getAdjPlot(mt,mt$disp,mt$mpg,mt$am,"Displacement","MPG","Transmission")
plotHP <- getAdjPlot(mt,mt$hp,mt$mpg,mt$am,"Gross Horespower","MPG","Transmission")
plotDRAT <- getAdjPlot(mt,mt$drat,mt$mpg,mt$am,"Rear Axle Ratio","MPG","Transmission")
```

## Adjustment Plot Observations (Adjustment Plot in Appendix)

1. Except horsepower all the other three viz. Weight, Displacement and Gear Axle Ratio explain for a lot of variation in MPG. 
2. We need to do here is a residual analysis. We need to study the how much of the residual variation in MPG is explained by Transmission after removing the effect of weight, hp, disp and drat one by one.

Before residual analysis let us do a quick ANOVA to see whether we should add no of cylinders in our model or not since it is too close to call there.

```{r anova}
#We have created a new dataset:mt out of mtcars where all categorical variables are converted to factors. 
#Hence explicit as.factor is not needed here.
fit1 <- lm(mpg~wt,data=mt)
fit2 <- update(fit1,mpg~wt+cyl)
fit3 <- update(fit1,mpg~wt*cyl)
anova(fit1,fit2,fit3)
```
## Anova Observation

Thus we can see that cylinder is significant here and we will add it to our residual analysis for weight. Let us add it for others too (but not its interaction as is evident from anova figures above).
```{r gettingResiduals}
mt$resid.mpg.wt <- resid(glm(mpg~wt+cyl,data=mt))  
mt$resid.am.wt <- resid(glm(am~wt+cyl,data=mt,family="binomial"))

mt$resid.mpg.disp <- resid(glm(mpg~disp+cyl,data=mt))  
mt$resid.am.disp <- resid(glm(am~disp+cyl,data=mt,family="binomial"))

mt$resid.mpg.hp <- resid(glm(mpg~hp+cyl,data=mt))  
mt$resid.am.hp <- resid(glm(am~hp+cyl,data=mt,family="binomial"))

mt$resid.mpg.drat <- resid(glm(mpg~drat+cyl,data=mt)) 
mt$resid.am.drat <- resid(glm(am~drat+cyl,data=mt,family="binomial"))
```

```{r residualanalysis,echo=FALSE,results='hide',fig.show='hide'}
getResidPlot <- function(df,xvar,yvar,xlab,ylab)
{
  residPlot <- ggplot(data=df, aes(x=xvar,y=yvar)) +
  geom_point(aes(color=abs(yvar),size=abs(yvar))) +  
  labs(title = "",
       subtitle = "",
       y = ylab, x = xlab) +
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(color = FALSE, size = FALSE) +
  scale_color_continuous(low = "black", high = "red") +
  geom_segment(aes(xend = xvar, yend = 0, color=), alpha = .2) + 
  geom_hline(yintercept=0)

  return(residPlot)
}

plotResidWeight <- getResidPlot(mt,mt$resid.am.wt,mt$resid.mpg.wt,"Residual (Transmission vs Weight)","Residual (MPG vs Weight)")
plotResidDisp <- getResidPlot(mt,mt$resid.am.disp,mt$resid.mpg.disp,"Residual (Transmission vs Displacement)","Residual (MPG vs Displacement)")
plotResidHP <- getResidPlot(mt,mt$resid.am.hp,mt$resid.mpg.hp,"Residual (Transmission vs HP)","Residual (MPG vs HP)")
plotResidDRAT <- getResidPlot(mt,mt$resid.am.drat,mt$resid.mpg.drat,"Residual (Transmission vs Gear Axle Ratio)","Residual (MPG vs Gear Axle Ratio)")
```

## Residual Plot Observations (Residual Plot in Appendix)
All the residual plots are homoscedastic. That is we cannot make out any pattern from these plots. Eg -  ***If the effect of weight and number of cylinders is removed, merely the fact that an engine is automatic or manual does not seem to have any impact on the MPG. The same holds true for effect of Displacement, HP and Gear Axle Ratio combined with No. of Cylinders***

## Question 2 is "Quantify the MPG difference between automatic and manual transmissions"

If we do not account for any other factors then it is the difference of means + or - the pooled Standard Deviation. Let us calculate that.

```{r getMeanDiff}
getTStats <- function(x1,x2,printLabel)
{
  s1 <- sd(x1); s2 = sd(x2); n1 <- length(x1); n2 <- length(x2);
  s <- sqrt(((n1-1)*(s1**2) + (n2-1)*(s2**2)) / (n1+n2-2))
  mu1 <- mean(x1); mu2 <- mean(x2)
  diffMU <- mu2 - mu1
  t.statistic <- diffMU / (s * sqrt((1/n1)+(1/n2)))
  p.value <- pt(t.statistic,n1+n2-2,lower.tail = FALSE)
  
  print(paste0("Difference in Means",printLabel,": ",round(diffMU,2)))
  print(paste0("Pooled SD",printLabel,": ",round(s,2)))

  return(p.value)
  
}
x1 <- mt[mt$am=="automatic",]$mpg; x2 <- mt[mt$am=="manual",]$mpg
x1.adj <- mt[mt$am=="automatic",]$resid.mpg.wt; x2.adj <- mt[mt$am=="manual",]$resid.mpg.wt

p.value <- getTStats(x1,x2,"")
p.value.adj <- getTStats(x1.adj,x2.adj," on Adjustment")
```

If we do not account for any other variable, the difference between means of MPG amongst manual and automatic transmission is significant to a p-value of **`r p.value`**. 

If we account for other variables like weight and no of cylinders, the difference between means of MPG amongst manual and automatic transmission is insignificant to a p-value of **`r p.value.adj`**. 

# Conclusion

1. On the face of it it does appear that Transmission is highly correlated with MPG
2. However there are other factors like weight, displacement, gear axle ratio and number of cylinders which also seem to impact MPG. They are also correlated with each other
3. When the effect of such other factors are adjusted along with number of cylinders for eg- weight + cylinders, we cannot come to a conclusion that there is any residual correlation between MPG and Transmission (i.e. Automatic or Manual)
4. **Hence Question #1 cannot be answered with the given data**
5. **Consequentially Question #2 implies that there is no statistically significant difference between MPG for Automatic and Manual Transmission Engines by mere virtue of their transmission type (after adjusting for other factors)**


# Appendix

## Box Plot
```{r, boxPlots, echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.height=3.5}
figure.box <- ggarrange(plotVS, plotCYL, plotGEAR, plotCARB, ncol = 2, nrow = 2)
figure.box
```

## Heat Map
```{r, heatMapDisplay, echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.height=3.5}
heatMap
```

## Adjustment Plots
```{r, adjustmentPlotDisplay, echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.height=3.5}
figure.point <- ggarrange(plotWeight, plotDisp, plotHP, plotDRAT, ncol = 2, nrow = 2)
figure.point
```


## Residual Plot
```{r, residualPlot, echo=FALSE, warning=FALSE, results='hide', message=FALSE, fig.height=3.5}
figure.residual <- ggarrange(plotResidWeight, plotResidDisp, plotResidHP, plotResidDRAT, ncol = 2, nrow = 2)
figure.residual
```
