---
title: "MT Cars Analysis"
author: "Jayesh Gokhale"
date: "5/1/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis on mtcars dataset

```{r loadLibraries, warning=FALSE, results='hide', message=FALSE}
library(datasets)
library(ggplot2)
library(RColorBrewer)
library(combinat)
library(knitr)
library(kableExtra)
```

```{r mutateData,echo=FALSE}
mt <- mtcars
mt$am <- as.factor(ifelse(mt$am==0,"automatic","manual"))
mt$vs <- as.factor(ifelse(mt$vs==0,"v-shaped","straight"))
mt$cyl <- as.factor(mt$cyl)
mt$gear <- as.factor(mt$gear)
mt$carb <- as.factor(mt$carb)
```

```{r basic plots}
ggplot(data=mt, aes(x=am,y=mpg,fill=cyl)) +
  geom_boxplot(outlier.colour="black", 
               outlier.size=2,position=position_dodge(1)) + 
  labs(title = "Motor Trend Car Road Tests",
       subtitle = "",
       y = "MPG", x = "Transmission") +
    scale_fill_brewer(name = "Cylinder",palette="Accent")  +
  theme(plot.title = element_text(hjust = 0.5)) 

ggplot(data=mt, aes(x=am,y=mpg,fill=vs)) +
  geom_boxplot(outlier.colour="black", 
               outlier.size=2,position=position_dodge(1)) + 
  labs(title = "Motor Trend Car Road Tests",
       subtitle = "",
       y = "MPG", x = "Transmission") +
    scale_fill_brewer(name = "Engine",palette="Accent")  +
  theme(plot.title = element_text(hjust = 0.5)) 

ggplot(data=mt, aes(x=am,y=mpg,fill=gear)) +
  geom_boxplot(outlier.colour="black", 
               outlier.size=2,position=position_dodge(1)) + 
  labs(title = "Motor Trend Car Road Tests",
       subtitle = "",
       y = "MPG", x = "Transmission") +
    scale_fill_brewer(name = "Gear",palette="Accent")  +
  theme(plot.title = element_text(hjust = 0.5)) 

ggplot(data=mt, aes(x=am,y=mpg,fill=carb)) +
  geom_boxplot(outlier.colour="black", 
               outlier.size=2,position=position_dodge(1)) + 
  labs(title = "Motor Trend Car Road Tests",
       subtitle = "",
       y = "MPG", x = "Transmission") +
    scale_fill_brewer(name = "Carburetor",palette="Accent")  +
  theme(plot.title = element_text(hjust = 0.5)) 
```
Let us plot basic scatter plot

```{r basicscatter}
plot(mt)
```

Let us build a model from numeric to numeric first
```{r}
lr.numeric <- glm(mpg~disp+hp+drat+wt+qsec,data=mt)
summary(lr.numeric)
```

From factors we will choose cylinder since as per plots, gear, carb, vs are not adding much value it seems

**This is clearly a trick question. The thing is that we have to find some other variable which are impacting MPG and Transmission is just a co-incidence. That is the crux of the problem**

**Similar to Breath mint**
